---
title: "Covid-19 data final"
author: "Vincent F"
date: "2025-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import Packages
Import required packages

```{r import}
library(tidyverse)
library(lubridate)
library(hms)
library(ggplot2)
library(dplyr)
```

## URLS for data sources

Get the data URLS

```{r urls}
urls <- c("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv",
            "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv",
            "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv",
            "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv"
            )
```

## read in the datasets into variable names
Create the dataframes

```{r variables}
global_cases <- read_csv(urls[2])
global_Deaths <- read_csv(urls[4])
us_cases <- read_csv(urls[1])
us_deaths <- read_csv(urls[3])
```
## fixing
cleaning and transforming
```{r cleaning}
#tidying global cases and deaths
global_cases <- global_cases %>%
  pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long),
               names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat,Long))

global_Deaths <- global_Deaths %>%
  pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long),
               names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat,Long))

global <- global_cases %>%
  full_join(global_Deaths) %>%
  rename(Country_Region = 'Country/Region',
          Province_State = 'Province/State') %>%
  mutate(date = mdy(date))

#removing global cases and deaths because I want to just focus on US cases for my analysis
rm(global_cases,global_Deaths)

#tidying the US DFs
us_cases <- us_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date=mdy(date))%>%
  select(-c(Lat,Long_))

us_deaths <- us_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date=mdy(date))%>%
  select(-c(Lat,Long_))

#join the us datasets
us<- us_cases %>%
  full_join(us_deaths)

#show summary of joined set
summary(us)
```
## grouping and sorting
Grouping the US by states
```{r grouping-by-state}
#group by state
us_by_state <- us %>%
  group_by(Province_State, Country_Region, date) %>%
    summarize(cases = sum(cases), deaths=sum(deaths),
              Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population)%>%
  select(Province_State, Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

```
## showing this on a graph based on latest date available
```{r us-by-state-graph}
# pick the most recent date in your data
latest_date <- max(us_by_state$date)


# filter for that date and sort by deaths_per_mill
df_latest <- us_by_state %>%
  filter(date == latest_date) %>%
  arrange(deaths_per_mill)

# plot a horizontal bar chart
ggplot(df_latest, aes(x = reorder(Province_State, deaths_per_mill),
                      y = deaths_per_mill)) +
  geom_col() +
  coord_flip() +
  labs(
    title = paste("COVID-19 Deaths per Million by State on", latest_date),
    x     = "State",
    y     = "Deaths per Million"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 6)  # shrink state labels if you have many
  )
```
## grouping and sorting
Grouping the US by totals
```{r grouping-by-totals}
#group by state
us_totals  <- us_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000/Population) %>%
  ungroup()

```
## plotting us_totals
```{r  us_totals_plot}
us_totals %>%
  filter(cases>0)%>%
  ggplot(aes(x=date,y=cases))+
  geom_line(aes(color="cases"))+
  geom_point(aes(color="cases"))+
  geom_line(aes(y=deaths,color="deaths"))+
  geom_point(aes(y=deaths,color="deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90))+
  labs(title="COVID19 in US", y = NULL)
```
##state of NY plot
```{r state-of-ny-plot}
state<- "New York"
us_by_state %>%
  filter(Province_State == state) %>%
  filter(cases>0)%>%
  ggplot(aes(x=date,y=cases))+
  geom_line(aes(color="cases"))+
  geom_point(aes(color="cases"))+
  geom_line(aes(y=deaths,color="deaths"))+
  geom_point(aes(y=deaths,color="deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90))+
  labs(title="COVID19 in NY", y = NULL)
```
## Analyzing...
```{r analyzing}
max(us_totals$deaths)
max(us_totals$date)

us_state_totals  <- us_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths=max(deaths), cases=max(cases),
            Population = max(Population),
            cases_per_thou=1000*cases/Population,
            deaths_per_thou=1000*deaths/Population) %>%
  filter(cases > 0, Population >0 )

us_state_totals %>% slice_min(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())

us_state_totals %>%
  slice_max(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())

```
## model data

```{r model}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = us_state_totals)
summary(mod)

us_state_totals %>% slice_min(cases_per_thou)
us_state_totals %>% slice_max(cases_per_thou)

us_state_totals %>% mutate(pred=predict(mod))

us_total_w_pred <- us_state_totals %>% mutate(pred=predict(mod))
```
## plot actual vs predictions
```{r plot-preds}
us_total_w_pred %>% ggplot() +
  geom_point(aes(x=cases_per_thou, y = deaths_per_thou), color = "blue")+
  geom_point(aes(x=cases_per_thou, y = pred), color = "red")
```